---
title: What Drives Engagement on YouTube?
subtitle: A Data Analysis of Influencer Metrics
author: Athina Agiakatsikas & Stella Christou
date: 2025-12-10
format:
  revealjs:
    theme: default
    transition: slide
    background-transition: fade
execute:
  warning: false
  message: false
  echo: false
---

# Introduction

## Research question

What factors most strongly influence engagement on YouTube?

### Examples of factors

-   Number of views
-   How long people watch
-   Day of the week posted
-   Number of likes and comments

## Why this matters

Understanding what drives engagement helps us:

-   Make better content
-   Post at the right time
-   Grow our audience faster

------------------------------------------------------------------------

# Data Overview

## What data did we analyze?

-   Time period: September 2019 - November 2024 (5 years)
-   Number of videos: About 300 videos
-   Daily records: 234,889 observations
-   Source: YouTube Analytics (official data)

## What we measured

We looked at:

-   Views: How many people watched
-   Likes: Approval
-   Comments: Active engagement
-   Watch time: How long people watched
-   Subscribers gained: New followers

------------------------------------------------------------------------

# What is Engagement?

## Simple definition

Engagement = How much people interact with your videos

$$\text{Engagement Rate} = \frac{\text{Likes + Comments}}{\text{Views}} \times 100$$

## How do we measure if the engagement rate is good or bad?

There is no universal standard for engagement across all youtube channels, so we used our own data

------------------------------------------------------------------------

# Our Findings

## First steps

**Status of our data set**: Messy, has gaps, duplicates, mixed-up data types

**Example of what "messy" looks like:**

```         
date          | views | likes | comments | watch_time | notes
2019-09-01    | 1500  | 45    | (blank)  | 12500      | (incomplete)
2019-09-01    | 1500  | 45    | (blank)  | 12500      | (duplicate row!)
2019-09-02    | NULL  | 50    | 3        | (text!)    | (broken data)
```

**What we did:** Clean up this mess into usable files

------------------------------------------------------------------------

```{r}
library(tidyverse)
video_perf <- read_csv("../data/video_performance.csv")

summary_stats <- video_perf %>%
  summarise(
    avg_engagement = paste0(round(mean(engagement_rate, na.rm = TRUE), 2), "%"),
    median_engagement = paste0(round(median(engagement_rate, na.rm = TRUE), 2), "%"),
    median_like_rate = paste0(round(median(like_rate, na.rm = TRUE), 2), "%"),
    median_comment_rate = paste0(round(median(comment_rate, na.rm = TRUE), 2), "%")
  )

knitr::kable(summary_stats)
```

## Key statistics

-   Average engagement: 3.32% (meaning on average, 3.32 out of 100 viewers like or comment)
-   Median engagement: 1.89% (middle value - many videos have lower engagement)
-   Likes are more common: About 1.76% average like rate
-   Comments are rare: Only 0.09% average comment rate (comments are 15-20x less common than likes!)

------------------------------------------------------------------------

# What Drives Engagement?

## The correlation analysis explained

Correlation = How two things move together

-   +1.0 = Perfect positive relationship (both go up together)
-   0 = No relationship
-   -1.0 = Perfect negative relationship (one goes up, other goes down)

We look for numbers close to +1.0 or -1.0 to find what matters!

## My results - ranked

```{r}
drivers <- data.frame(
  Factor = c("Like Rate", "View Percentage", "Comment Rate", "Watch Duration"),
  Correlation = c("0.99", "0.37", "0.35", "-0.51"),
  Meaning = c(
    "Strongest predictor! Likes show engagement",
    "Videos people watch fully get more engagement",
    "People who comment are engaged",
    "Surprisingly: longer videos have LOWER engagement!"
  )
)

knitr::kable(drivers)
```

------------------------------------------------------------------------

# Key Insight 1: Quality Beats Quantity

Videos with fewer views can have HIGHER engagement rates!

-   A video with 1,000 views and 50 likes = 5% engagement (great!)
-   A video with 100,000 views and 500 likes = 0.5% engagement (weak)

Lesson: Focus on making content your audience cares about, not just getting views.

------------------------------------------------------------------------

# Key Insight 2: Best Days to Post

```{r}
weekly <- read_csv("../data/weekly_pattern.csv")

ggplot(weekly, aes(x = fct_inorder(day_of_week), y = avg_engagement_rate, fill = day_of_week)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Average Engagement by Day of Week",
    x = "Day of Week",
    y = "Engagement Rate (%)"
  ) +
  theme_minimal()
```

Look at the chart and find the tallest bar - that's your best day to post!

------------------------------------------------------------------------

# Key Insight 3: Engagement Trends Over Time

```{r}
daily <- read_csv("../data/daily_performance.csv")

ggplot(daily, aes(x = date, y = avg_engagement_rate)) +
  geom_line(color = "#2E86AB") +
  geom_smooth(method = "loess", color = "#A23B72", alpha = 0.2) +
  labs(
    title = "Engagement Rate Trend Over Time",
    x = "Date",
    y = "Average Engagement Rate (%)"
  ) +
  scale_x_date(date_breaks = "6 months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

------------------------------------------------------------------------

# Key Insight 4: Likes vs Comments

```{r}
engagement_types <- data.frame(
  Type = c("Likes", "Comments"),
  Percentage = c(
    round(median(video_perf$like_rate, na.rm = TRUE), 2),
    round(median(video_perf$comment_rate, na.rm = TRUE), 2)
  )
)

ggplot(engagement_types, aes(x = Type, y = Percentage, fill = Type)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Median: Likes vs Comments", y = "Percentage of Viewers (%)") +
  theme_minimal()
```

## The takeaway

-   Likes: Easy to give, common (1.76% median)
-   Comments: Harder to give, rare (0.09% median), but show REAL engagement
-   Ratio: About 15-20 likes for every 1 comment

Comments are gold! They show people really care about your content.

------------------------------------------------------------------------

# My Recommendations

## 1. Focus on content quality

Not views - engagement!

-   Make videos your audience genuinely cares about
-   A video with 1,000 engaged viewers is better than 100,000 bored viewers
-   Check your best-performing videos and see what they have in common

## 2. Encourage comments

Comments show real engagement

-   Ask questions in your videos
-   Respond to comments (builds community)
-   Comments = free advertising (YouTube shows them to others)

## 3. Post at the right time

Use the day-of-week chart

-   Post when your audience is active
-   Different days work for different audiences
-   Test and monitor what works for YOU

## 4. Watch retention matters

Keep people watching

-   Shorter, focused videos can outperform long ones
-   Hook viewers in the first 10 seconds
-   Get to the point quickly

## 5. Track your progress

Monitor these metrics monthly

-   Engagement rate (likes + comments รท views)
-   Like rate
-   Comment rate
-   Subscriber growth

------------------------------------------------------------------------

# Summary

## What we learned

1.  Quality \> Quantity: High engagement beats high view counts
2.  Timing matters: Post when your audience is online
3.  Comments are valuable: They show real interest
4.  Engagement drives growth: More engagement = more subscribers
5.  Watch retention is key: Keep people watching until the end

## Your numbers

-   Average engagement: 3.32%
-   Best interaction type: Likes (15-20x more common than comments)
-   Best day to post: Check your data (weekly_pattern.csv)

## Next steps

1.  Apply these insights to your content strategy
2.  Track your metrics monthly
3.  Test different posting times and content types
4.  Monitor which videos get the most engagement

------------------------------------------------------------------------

# Questions?
