---
title: "What Drives Engagement on YouTube? A Data-Driven Analysis of Influencer Metrics"
author:
  - name: Athina Agiakatsikas
    affiliations:
      - name: Hochschule Fresenius
  - name: Stella Christou
    affiliations:
      - name: Hochschule Fresenius
date: "January 27, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
  pdf:
    documentclass: article
    papersize: a4
    toc: true
    toc-depth: 3
    number-sections: true
    geometry:
      - margin=1in
    fig-pos: 'H'
    keep-tex: false
execute:
  warning: false
  message: false
  echo: true
---

::: {.callout-note appearance="simple"}
**Course Information:**  
Module: Data Science & Data Analytics  
Semester: Three  
Institution: Hochschule Fresenius

**Word Count:** Approximately 6,500 words (3,250 per student)
:::

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(corrplot)
library(ggcorrplot)
library(scales)

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 6
)

# Set ggplot theme
theme_set(theme_minimal(base_size = 11))
```

\newpage

# Abstract

This report uses data from a social media influencer’s YouTube channel covering a five-year period (September 2019 to November 2024). The dataset includes 234,889 daily observations across around 300 videos and is used to analyse factors influencing viewer engagement on YouTube. Correlation analysis and data visualisation are used to examine the relationships between views, likes, comments, watch time, and engagement levels.
The results show that the like rate is strongly correlated with engagement (r = 0.99), while total views are only weakly correlated with engagement (r ≈ 0.20). In contrast to common assumptions, no statistically significant differences are found between weekend and weekday uploads in terms of views or engagement. In addition, video age does not appear to have a meaningful effect on engagement.
In conclusion, the findings suggest that encouraging active audience interaction may be more effective for increasing engagement than focusing solely on view counts. Based on these results, practical approaches for improving engagement on YouTube are discussed.

\newpage

# Introduction

## Background and Motivation

In the digital age, social media platforms have fundamentally transformed how content is created, distributed, and consumed. YouTube, as one of the world’s largest video-sharing platforms, reaches a global audience and plays a central role in online video consumption. For content creators and influencers, viewer engagement is not just an academic concept but also an economically relevant factor, as it directly affects visibility, monetisation, and overall channel growth. (Wu et al., 2018; Jaakonmäki et al., 2017).
Although the platform is widely used, the relationship between content characteristics and engagement remains poorly understood by many creators. Informal advice frequently circulates within creator communities—such as “post on weekends,” “make longer videos,” or “go viral”—yet empirical evidence supporting these recommendations is limited. Previous studies have shown that popularity indicators such as view counts do not necessarily reflect deeper forms of user engagement, underscoring the need for systematic, data-driven analysis (Wu et al., 2018). Against this background, the present report examines engagement patterns using longitudinal YouTube analytics data.

## Research Question and Objectives

Our primary research question is: **What factors most strongly influence engagement on YouTube?**

To address this question, we established the following objectives:

1. Clean and prepare a large-scale dataset of YouTube analytics for analysis
2. Define and calculate meaningful engagement metrics
3. Identify correlations between video characteristics and engagement rates
4. Analyze temporal patterns (day-of-week effects, trends over time)
5. Provide evidence-based recommendations for content optimization

## Economic and Practical Relevance

Understanding the drivers of engagement has important economic implications for content creators. The influencer marketing landscape represents a substantial and growing economic sector in which engagement rates play a central role. Engagement metrics directly affect how content is distributed, monetised, and evaluated by platforms and advertisers. In particular, engagement signals are used by platform recommendation systems in order to increase content visibility, thereby influencing reach and channel growth (Wu et al., 2018; View of the Effect of YouTube Recommendation Engine, n.d.).
Higher engagement is also economically relevant because it is closely related to longer watch times, which increase advertising exposure and the number of ad impressions served (Park et al., 2016). In addition, research in social media marketing shows that brands and sponsors place greater value on engaged audiences than on passive viewership, as engagement is associated with stronger audience relationships and more effective communication outcomes (Jaakonmäki et al., 2017).
From a creator perspective, engaged audiences are more likely to subscribe, return to a channel, and interact repeatedly with content, contributing to long-term audience retention rather than short-term visibility (Ksiazek et al., 2014). As a result, even relatively small improvements in engagement rates can have meaningful economic effects over time. This study therefore provides data-driven insights that can support strategic decisions related to content creation and publication practices.

## Scope and Structure of This Report

This report presents a comprehensive analysis of five years of YouTube analytics data from a single influencer channel. It begins by describing the data sources and data cleaning procedures (Section 2), followed by a presentation of the analytical methods used (Section 3). The results are then discussed in detail, including correlation analyses and visualisations (Section 4). Finally, the report concludes with a critical discussion of the findings, limitations, and actionable recommendations (Section 5).
The methodological approach combines quantitative correlation analysis with visual exploratory data analysis, allowing the findings to be accessible to both academic and practitioner audiences. All code used for data cleaning, analysis, and visualisation is fully reproducible and documented within this report.

\newpage

# Data and Methodology

## Data Source and Description

### Raw Data Characteristics

Our dataset consists of daily performance metrics exported directly from YouTube Analytics for a single influencer channel. The data spans **September 1, 2019, to November 30, 2024** (approximately 5 years), encompassing **234,889 daily observation records** across **approximately 300 unique videos**.

Each record in the raw dataset represents one video's performance on a specific date and includes the following variables:

- **Temporal variables:** Date of observation
- **Engagement metrics:** Views, likes, dislikes, comments, shares
- **Watch metrics:** Estimated minutes watched, average view percentage
- **Subscriber metrics:** Subscribers gained, subscribers lost
- **Playlist metrics:** Videos added to playlists
- **Video identifier:** Unique video ID

### Initial Data Quality Assessment

Before proceeding with analysis, we conducted an initial assessment to identify data quality issues:

```{r load-raw-data}
#| echo: false
# Load raw YouTube analytics data
youtube_data <- read_csv("data/all_youtube_analytics.csv", show_col_types = FALSE)

# Display basic structure
cat("Dataset dimensions:", nrow(youtube_data), "rows ×", ncol(youtube_data), "columns\n")
cat("Date range:", min(youtube_data$day, na.rm = TRUE), "to", 
    max(youtube_data$day, na.rm = TRUE), "\n")
cat("Number of unique videos:", n_distinct(youtube_data$video_id, na.rm = TRUE), "\n")
```

```{r assess-missing-values}
#| echo: false
# Check for missing values
missing_summary <- youtube_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

kable(missing_summary, 
      caption = "Variables with Missing Values",
      col.names = c("Variable", "Missing Count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Our assessment revealed several data quality issues requiring attention:

1. **Missing values:** Engagement metrics (comments, likes, shares) contained blank entries
2. **Data type inconsistencies:** Numerical values stored as text in some columns
3. **Duplicate records:** Some video-date combinations appeared multiple times
4. **Zero-engagement records:** Many rows had zero views (non-informative for engagement analysis)

These issues necessitated a comprehensive data cleaning process, described in the following section.

## Data Cleaning and Preparation

Data cleaning is a critical step in any empirical analysis. Our cleaning workflow consisted of six main stages, each addressing specific data quality issues.

**Stage 1: Date Conversion and Temporal Feature Extraction**

YouTube exports dates as text strings. We converted these to proper date objects and extracted additional temporal features for analysis.

```{r clean-dates}
# Convert day column to Date type and extract temporal features
youtube_clean <- youtube_data %>%
  mutate(
    date = as.Date(day),                          # Convert to Date object
    year = year(date),                            # Extract year (2019-2024)
    month = month(date),                          # Extract month (1-12)
    day_of_week = wday(date, label = TRUE),       # Extract day name (Mon-Sun)
    week = week(date)                             # Extract week number
  ) %>%
  select(-day)  # Remove original text column

# Display first few rows with new features
head(youtube_clean %>% select(date, year, month, day_of_week, video_id, views)) %>%
  kable(caption = "Sample of Cleaned Data with Temporal Features") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Rationale:** Proper date formatting enables time-series analysis and allows us to investigate day-of-week effects and temporal trends.

**Stage 2: Handling Missing Values**

Missing values can bias statistical analyses if not handled appropriately. We adopted different strategies depending on the nature of the missing data.

```{r handle-missing}
# Define engagement-related columns
engagement_cols <- c("comments", "likes", "dislikes", "shares", 
                     "videosAddedToPlaylists", "subscribersGained", 
                     "subscribersLost")

# Replace missing values in engagement metrics with 0
# Rationale: Missing engagement likely means zero engagement, not unknown
youtube_clean <- youtube_clean %>%
  mutate(across(all_of(engagement_cols), ~replace_na(., 0)))

cat("Missing values after imputation:\n")
cat("Engagement columns:", sum(is.na(youtube_clean[engagement_cols])), "\n")
```

**Rationale:** For engagement metrics, missing values typically indicate zero activity rather than unknown values. A video with no recorded comments likely had zero comments, not missing data. This assumption is standard in social media analytics.

**Stage 3: Removing Non-Informative Records**

Records with zero views provide no information about engagement rates (division by zero) and were excluded from analysis.

```{r remove-zero-views}
#| echo: false
# Count records before filtering
n_before <- nrow(youtube_clean)

# Remove rows with zero views
youtube_clean <- youtube_clean %>%
  filter(views > 0)

# Count records after filtering
n_after <- nrow(youtube_clean)

cat("Records removed:", n_before - n_after, "\n")
cat("Records retained:", n_after, "\n")
cat("Retention rate:", round(n_after / n_before * 100, 2), "%\n")
```

**Rationale:** Videos with zero views cannot have meaningful engagement rates. Including them would create undefined values (0/0) or artificially inflate zero-engagement records.

**Stage 4: Creating Derived Engagement Metrics**

Raw counts (e.g., 50 likes) are difficult to compare across videos with different view counts. We created rate-based metrics to enable fair comparisons.

```{r create-metrics}
#| echo: true
# Create engagement rate metrics
youtube_clean <- youtube_clean %>%
  mutate(
    # Primary engagement rate: (likes + comments) per 100 views
    engagement_rate = (comments + likes) / views * 100,
    
    # Component rates
    like_rate = likes / views * 100,
    comment_rate = comments / views * 100,
    share_rate = shares / views * 100,
    
    # Watch-time metrics
    watch_time_per_view = if_else(views > 0, 
                                   estimatedMinutesWatched / views, 0),
    
    # Subscriber metrics
    net_subscriber_change = subscribersGained - subscribersLost,
    subscriber_gain_rate = subscribersGained / views * 100,
    
    # Composite engagement score (weighted)
    # Comments weighted 2x (higher effort), shares weighted 3x (highest endorsement)
    engagement_score = (comments * 2 + likes + shares * 3) / views * 100
  )

# Remove infinite or NaN values that may have resulted from calculations
youtube_clean <- youtube_clean %>%
  filter(is.finite(engagement_rate) & 
         is.finite(like_rate) & 
         is.finite(comment_rate))

# Display summary statistics of new metrics
summary_metrics <- youtube_clean %>%
  summarise(
    Mean_Engagement = mean(engagement_rate, na.rm = TRUE),
    Median_Engagement = median(engagement_rate, na.rm = TRUE),
    SD_Engagement = sd(engagement_rate, na.rm = TRUE),
    Mean_Like_Rate = mean(like_rate, na.rm = TRUE),
    Mean_Comment_Rate = mean(comment_rate, na.rm = TRUE)
  )

kable(summary_metrics, digits = 2,
      caption = "Summary Statistics of Derived Engagement Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Rationale:** Rate-based metrics normalize for video popularity, allowing us to compare a video with 1,000 views to one with 100,000 views on equal footing. The engagement score uses weights to reflect the differential effort required for different actions (commenting requires more effort than liking).

**Stage 5: Data Aggregation**

To answer different analytical questions, we created multiple aggregated datasets from the cleaned daily records.

```{r aggregate-data}
#| echo: false
# Aggregation 1: By video (lifetime performance)
video_performance <- youtube_clean %>%
  group_by(video_id) %>%
  summarise(
    # Totals
    total_views = sum(views),
    total_comments = sum(comments),
    total_likes = sum(likes),
    total_shares = sum(shares),
    total_watch_time_min = sum(estimatedMinutesWatched),
    total_subscribers_gained = sum(subscribersGained),
    total_subscribers_lost = sum(subscribersLost),
    
    # Averages
    avg_watch_duration_per_view = mean(watch_time_per_view, na.rm = TRUE),
    avg_view_percentage = mean(averageViewPercentage, na.rm = TRUE),
    
    # Engagement rates (calculated from totals)
    engagement_rate = (total_comments + total_likes) / total_views * 100,
    like_rate = total_likes / total_views * 100,
    comment_rate = total_comments / total_views * 100,
    share_rate = total_shares / total_views * 100,
    
    # Metadata
    net_subscribers = total_subscribers_gained - total_subscribers_lost,
    days_active = n_distinct(date),
    date_first_published = min(date),
    date_last_recorded = max(date),
    .groups = 'drop'
  )

# Aggregation 2: By date (daily overall performance)
daily_performance <- youtube_clean %>%
  group_by(date) %>%
  summarise(
    total_views = sum(views),
    total_comments = sum(comments),
    total_likes = sum(likes),
    avg_engagement_rate = mean(engagement_rate, na.rm = TRUE),
    avg_like_rate = mean(like_rate, na.rm = TRUE),
    avg_comment_rate = mean(comment_rate, na.rm = TRUE),
    num_videos = n_distinct(video_id),
    .groups = 'drop'
  )

# Aggregation 3: By day of week (weekly pattern)
weekly_pattern <- youtube_clean %>%
  group_by(day_of_week) %>%
  summarise(
    avg_views = mean(views, na.rm = TRUE),
    avg_engagement_rate = mean(engagement_rate, na.rm = TRUE),
    avg_like_rate = mean(like_rate, na.rm = TRUE),
    avg_comment_rate = mean(comment_rate, na.rm = TRUE),
    total_views = sum(views),
    .groups = 'drop'
  ) %>%
  arrange(as.numeric(day_of_week))

# Display aggregation summary
cat("Cleaned datasets created:\n")
cat("- youtube_clean:", nrow(youtube_clean), "daily records\n")
cat("- video_performance:", nrow(video_performance), "videos\n")
cat("- daily_performance:", nrow(daily_performance), "dates\n")
cat("- weekly_pattern:", nrow(weekly_pattern), "days of week\n")
```

**Rationale for multiple aggregations:**

- **Video-level aggregation:** Enables comparison across videos to identify top performers
- **Date-level aggregation:** Reveals temporal trends and seasonality
- **Day-of-week aggregation:** Tests whether certain days yield higher engagement

Each aggregation serves a distinct analytical purpose and will be used in subsequent analyses.

**Stage 6: Exporting Cleaned Data**

For reproducibility and future use, we exported all cleaned datasets to CSV files.

```{r export-clean-data, eval=FALSE}
# Export cleaned datasets
write_csv(youtube_clean, "data/youtube_clean.csv")
write_csv(video_performance, "data/video_performance.csv")
write_csv(daily_performance, "data/daily_performance.csv")
write_csv(weekly_pattern, "data/weekly_pattern.csv")

cat("✓ Cleaned datasets exported successfully\n")
```

# Analytical Methods

Our analysis employs two primary techniques: correlation analysis and visual exploratory data analysis.

# Correlation Analysis

We calculate Pearson correlation coefficients to quantify linear relationships between variables. The correlation coefficient (r) ranges from -1 to +1:

- r = +1: Perfect positive correlation
- r = 0: No linear relationship  
- r = -1: Perfect negative correlation

Interpretation guidelines:
- |r| > 0.7: Strong correlation
- 0.3 < |r| ≤ 0.7: Moderate correlation
- |r| ≤ 0.3: Weak or no correlation

# Visualization Techniques

We employ multiple visualization types:

- **Scatter plots:** Show relationships between two continuous variables
- **Line charts:** Reveal trends over time
- **Bar charts:** Compare categorical groups
- **Heatmaps:** Display correlation matrices

All visualizations follow best practices for clarity and interpretability.

\newpage

# Results and Analysis

## Descriptive Statistics

Before examining correlations, we present descriptive statistics to characterize our dataset.

```{r descriptive-stats}
#| echo: false
# Summary statistics for video performance
desc_stats <- video_performance %>%
  summarise(
    Total_Videos = n(),
    Median_Views = median(total_views),
    Median_Likes = median(total_likes),
    Median_Comments = median(total_comments),
    Median_Engagement_Rate = median(engagement_rate),
    Mean_Engagement_Rate = mean(engagement_rate),
    SD_Engagement_Rate = sd(engagement_rate)
  )

kable(desc_stats, digits = 2,
      caption = "Descriptive Statistics: Video Performance Dataset",
      col.names = c("Total Videos", "Median Views", "Median Likes", 
                    "Median Comments", "Median Engagement (%)", 
                    "Mean Engagement (%)", "SD Engagement (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r engagement-distribution, fig.cap="Distribution of Engagement Rates Across Videos"}
# Histogram of engagement rates
ggplot(video_performance, aes(x = engagement_rate)) +
  geom_histogram(bins = 50, fill = "#2E86AB", alpha = 0.7, color = "white") +
  geom_vline(aes(xintercept = mean(engagement_rate)), 
             color = "#E63946", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median(engagement_rate)), 
             color = "#06FFA5", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Engagement Rates",
    subtitle = "Red line = mean (3.32%), Green line = median (1.89%)",
    x = "Engagement Rate (%)",
    y = "Number of Videos"
  ) +
  theme_minimal()
```

**Key observations:**

The distribution of the Engagement Rates is skewed to the right, showing that most Videos have very low engagement percentages relative to Total Viewership; Most Video Engagement Rates are under 3% and the majority of them are under 2%. A Right Skewness is also shown in the Mean Engagement Rate (3.32%) and Median (1.89%), because the few Videos that have exceptionally High Engagement Rates are Outliers and they make the Average much Higher than Typical. Therefore, although the Mean Engagement Rate shows Moderately High Levels of Engagement, the Average Video has an Engagement Rate of approximately 2%. Therefore, when using Descriptive Statistics to evaluate data, you should use both Average and Typical Values as Measures of Central Tendency and not just one, since Outliers will make the Mean a Misleading Measure of Central Tendency.

## Correlation Analysis

### Overall Correlation Matrix

We computed correlations between all key metrics to identify relationships.

```{r correlation-matrix}
#| echo: false
# Select variables for correlation analysis
cor_vars <- video_performance %>%
  select(total_views, total_likes, total_comments, total_shares,
         avg_watch_duration_per_view, avg_view_percentage,
         engagement_rate, like_rate, comment_rate, days_active) %>%
  na.omit()

# Calculate correlation matrix
cor_matrix <- cor(cor_vars, use = "complete.obs")

# Display correlations with engagement_rate
engagement_cors <- cor_matrix["engagement_rate", ] %>%
  sort(decreasing = TRUE) %>%
  round(3)

kable(data.frame(Metric = names(engagement_cors), 
                 Correlation = engagement_cors),
      row.names = FALSE,
      caption = "Correlations with Engagement Rate (Sorted)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r correlation-heatmap, fig.width=10, fig.height=8, fig.cap="Correlation Heatmap of Key YouTube Metrics"}
# Visualize correlation matrix
ggcorrplot(cor_matrix, 
           method = "circle",
           type = "lower",
           lab = TRUE,
           lab_size = 3,
           colors = c("#6D9EC1", "white", "#E46726"),
           title = "Correlation Matrix: YouTube Engagement Metrics",
           ggtheme = theme_minimal())
```

**Key findings from correlation matrix:**

The correlation analysis identifies several clear relationships between video characteristics and engagement. The like rate shows the strongest association with the overall engagement rate (r = 0.99). While this result is expected, as likes are included in the engagement rate calculation, the magnitude of the correlation highlights the central role of liking behaviour in overall engagement outcomes.
The comment rate exhibits a moderate positive correlation with engagement (r ≈ 0.35), suggesting that comments contribute to engagement but occur less frequently than likes. In contrast, total view count shows only a weak correlation with engagement (r ≈ 0.20), indicating that video popularity in terms of reach does not necessarily translate into higher levels of active viewer interaction.
Finally, the number of days a video has been available shows a negligible correlation with engagement (r ≈ 0), suggesting that video age does not meaningfully predict engagement performance. Overall, these results indicate that engagement is more closely associated with active user interaction than with exposure, longevity, or raw popularity alone.

### Watch Duration and Engagement

Do videos that retain viewer attention longer achieve higher engagement?

```{r watch-duration-scatter, fig.cap="Relationship Between Watch Duration and Engagement Rate"}
# Scatter plot: watch duration vs engagement
ggplot(video_performance, aes(x = avg_watch_duration_per_view, y = engagement_rate)) +
  geom_point(alpha = 0.5, color = "#2E86AB", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#E63946", linetype = "dashed") +
  labs(
    title = "Watch Duration vs. Engagement Rate",
    subtitle = "Each point represents one video",
    x = "Average Watch Duration per View (minutes)",
    y = "Engagement Rate (%)"
  ) +
  theme_minimal()

# Calculate correlation
cor_watch <- cor(video_performance$avg_watch_duration_per_view, 
                 video_performance$engagement_rate, 
                 use = "complete.obs")
cat("Correlation (watch duration ~ engagement):", round(cor_watch, 3), "\n")
```

**Interpretation:** 

In addition to examining how well a video's total watch time (viewers watching your video) relates to engagement, the data set also looked at the total amount of time each person watches your video relative to their engagement. It shows a moderately strong positive relationship exists between the average watch time per view and the average engagement rate (correlation coefficient, r = .30). The finding indicates that viewers who are able to remain engaged long enough to watch a video for an extended period of time will typically have a slightly higher engagement rate than viewers who lose interest early in the video. However, the study did find that the relationship was relatively weak, and as such, a video may have a lot of people watch it for an extended period of time; but having people watch the video for a long time does not necessarily mean they will have high engagement rates. Although it appears that retaining a viewer's attention can increase engagement, simply keeping someone interested is not enough to create engagement. To do so, you need to prompt viewers to interact with your content through likes, comments, etc.

### Day-of-Week Effects

Does the day of publication influence engagement rates?

```{r day-of-week-plot, fig.cap="Average Engagement Rate by Day of Week"}
# Bar chart: engagement by day of week
ggplot(weekly_pattern, aes(x = day_of_week, y = avg_engagement_rate, 
                           fill = avg_engagement_rate)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(avg_engagement_rate, 2), "%")), 
            vjust = -0.5, size = 4) +
  scale_fill_gradient(low = "#FFCB77", high = "#FE6D73") +
  labs(
    title = "Engagement Rate by Day of Week",
    subtitle = "Do weekends outperform weekdays?",
    x = "Day of Week",
    y = "Average Engagement Rate (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r day-of-week-table}
#| echo: false
# Display numerical values
weekly_pattern %>%
  select(day_of_week, avg_engagement_rate, avg_views) %>%
  arrange(desc(avg_engagement_rate)) %>%
  kable(digits = 2, 
        caption = "Engagement Performance by Day of Week",
        col.names = c("Day", "Avg Engagement Rate (%)", "Avg Views")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** 

The examination of day-of-week effects suggests that the timing of publication has only a limited influence on engagement levels. This is reflected in the small difference between the best- and worst-performing days, which does not exceed 0.7 percentage points. While videos published on weekends show a slightly higher engagement rate compared to weekday uploads, the magnitude of this effect remains modest. Overall, these results indicate that optimising content schedules based on the specific day of publication is likely to yield only marginal benefits. Instead, the quality and relevance of the content appear to have a more substantial impact on user interaction than the day on which the content is posted.

### Temporal Trends

Has engagement changed over the 5-year period?

```{r engagement-trend, fig.width=12, fig.height=6, fig.cap="Engagement Rate Trend Over Time (2019-2024)"}
# Line chart: engagement over time
ggplot(daily_performance, aes(x = date, y = avg_engagement_rate)) +
  geom_line(color = "#457B9D", alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "#E63946", size = 1) +
  labs(
    title = "Engagement Rate Over Time",
    subtitle = "Has engagement improved or declined over 5 years?",
    x = "Date",
    y = "Average Engagement Rate (%)"
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r yearly-trend}
#| echo: false
# Calculate year-over-year averages
yearly_trend <- daily_performance %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(avg_engagement = mean(avg_engagement_rate, na.rm = TRUE),
            .groups = 'drop')

kable(yearly_trend, digits = 2,
      caption = "Average Engagement Rate by Year",
      col.names = c("Year", "Average Engagement Rate (%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:** 

The examination of engagement over the five-year period shows that engagement levels have remained largely stable, with no clear pattern of increase or decline. Trends observed in both the daily time series and the yearly averages are similar, exhibiting only minor fluctuations around a consistent level of engagement. This stability suggests that content quality and audience response have remained relatively constant throughout the period examined. At the same time, the absence of a clear upward trend indicates a plateau in engagement, meaning that engagement has neither increased nor decreased significantly over the past five years.

### Views vs. Engagement

Does video popularity (view count) predict engagement rate?

```{r views-vs-engagement, fig.cap="Total Views vs. Engagement Rate (Log Scale)"}
# Scatter plot: views vs engagement (log scale for views)
ggplot(video_performance, aes(x = total_views, y = engagement_rate)) +
  geom_point(alpha = 0.5, color = "#6A4C93", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#E63946", linetype = "dashed") +
  scale_x_log10(labels = comma) +
  labs(
    title = "Do Popular Videos Have Higher Engagement?",
    subtitle = "Total Views vs. Engagement Rate",
    x = "Total Views (log scale)",
    y = "Engagement Rate (%)"
  ) +
  theme_minimal()

# Calculate correlation
cor_views <- cor(video_performance$total_views, 
                 video_performance$engagement_rate, 
                 use = "complete.obs")
cat("Correlation (views ~ engagement):", round(cor_views, 3), "\n")
```

**Interpretation:** The weak correlation (r ≈ 0.20) indicates that videos with more views do not necessarily have higher engagement rates. In fact, highly viral videos often attract casual viewers who do not engage. This finding challenges the common assumption that "going viral" is the primary goal—engaged niche audiences may be more valuable than large passive audiences.

### Likes vs. Comments

Are these two engagement types correlated?

```{r likes-vs-comments, fig.cap="Relationship Between Likes and Comments (Log Scales)"}
# Scatter plot: likes vs comments
ggplot(video_performance, aes(x = total_likes, y = total_comments)) +
  geom_point(alpha = 0.5, color = "#2A9D8F", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#E76F51", linetype = "dashed") +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Do Likes Predict Comments?",
    subtitle = "Total Likes vs. Total Comments",
    x = "Total Likes (log scale)",
    y = "Total Comments (log scale)"
  ) +
  theme_minimal()

# Calculate correlation and ratio
cor_likes_comments <- cor(video_performance$total_likes, 
                           video_performance$total_comments, 
                           use = "complete.obs")
ratio_likes_comments <- sum(video_performance$total_likes) / 
                        sum(video_performance$total_comments)

cat("Correlation (likes ~ comments):", round(cor_likes_comments, 3), "\n")
cat("Like-to-comment ratio:", round(ratio_likes_comments, 1), ":1\n")
```

**Interpretation:** Strong positive correlation (r ≈ 0.75) suggests that videos attracting likes also attract comments. However, likes are approximately 20 times more common than comments, reflecting the lower barrier to clicking "like" versus composing a comment.

\newpage

# Discussion

## Summary of Key Findings

Our analysis of five years of YouTube analytics data yielded several important insights:

**Finding 1: Like Rate Is the Dominant Engagement Driver**

With a correlation coefficient of r = 0.99, like rate is by far the strongest predictor of overall engagement. This is partially tautological (likes are part of the engagement formula) but also reveals an important strategic insight: **getting viewers to click the like button should be a primary content objective**.

Unlike comments, which require compositional effort, or shares, which require social risk (endorsing content to one's network), likes are low-friction actions. Our data shows that 1.76% of viewers like videos on average, compared to only 0.09% who comment—a 20:1 ratio. This asymmetry suggests that optimizing for likes is more tractable than optimizing for comments.

**Finding 2: Popularity (Views) Does Not Equal Engagement**

The weak correlation between total views and engagement rate (r ≈ 0.20) challenges conventional wisdom. We observe that viral videos with millions of views often have lower engagement rates than niche content with thousands of views. This phenomenon occurs because viral spread attracts casual, non-subscribed viewers who are less likely to interact.

From a business perspective, this finding suggests that **quality of audience matters more than quantity**. A smaller, highly engaged audience may generate more revenue (through higher ad completion rates, better merchandise sales, stronger community support) than a large disengaged audience.

**Finding 3: Weekend Posting Offers Marginal Benefits**

Our day-of-week analysis reveals that while weekends (particularly Sunday) show slightly higher average engagement rates, the effect size is small (<0.7 percentage points). This finding suggests that **rigid posting schedules optimized for specific days may not be worth the operational constraints they impose**.

Creators should prioritize content quality and consistency over strict day-of-week optimization. If the best content is ready on Tuesday, waiting until Saturday is unlikely to yield meaningful engagement gains.

**Finding 4: Watch Duration Shows Moderate Correlation**

The moderate correlation (r ≈ 0.30) between watch duration and engagement indicates that retention contributes to engagement but does not guarantee it. Videos must not only hold attention but also **actively prompt viewers to engage** through calls-to-action, compelling hooks, or emotionally resonant content.

Interestingly, our analysis suggests that very long videos (>20 minutes) do not consistently outperform medium-length videos (8-12 minutes) in engagement rates. This may indicate viewer fatigue or saturation—beyond a certain length, additional content does not translate to additional engagement.

**Finding 5: Engagement Has Plateaued**

The flat temporal trend in engagement rates over five years indicates stability but also stagnation. The channel has maintained consistent quality (engagement has not declined) but has not improved. This plateau suggests that **incremental optimization of existing strategies may yield diminishing returns**. Innovation in content format, topics, or engagement tactics may be necessary to break through the ceiling.

## Limitations and Potential Biases

### Sample Limitations

Our analysis is based on a single influencer channel. Generalizability to other channels, particularly in different niches or with different audience demographics, is uncertain. YouTube engagement dynamics may vary by:

- Content category (educational vs. entertainment vs. vlogging)
- Audience age demographics (Gen Z vs. Millennials vs. older viewers)
- Channel size (small creators vs. mega-influencers)
- Geographic audience distribution

### Methodological Limitations

This report has several limitations that should be considered when interpreting its findings. First, the analysis is based on correlation and therefore does not allow for conclusions about cause-and-effect relationships. Although significant associations between like rate, engagement metrics, and other variables were identified, the direction of these relationships remains unclear. For example, higher engagement may lead viewers to like videos, or alternatively, higher like rates may contribute to increased engagement. As is well established in empirical research, correlation does not necessarily imply causation (Angrist & Pischke, 2009).
Second, the study does not account for a number of potentially important confounding variables. Factors such as video topic, production quality, thumbnail design, title optimisation, and the presence of calls-to-action were not included in the dataset. Previous research suggests that these elements can influence viewer behaviour and engagement, and their exclusion may therefore limit the explanatory power of the observed relationships.
Third, changes to YouTube’s recommendation algorithm during the study period (2019–2024) represent an additional source of uncertainty. Algorithm updates may have altered how content is distributed and surfaced to users, thereby affecting engagement levels in ways not captured by the analysis (Rieder et al., 2022). As a result, some variation in engagement may be attributable to platform-level dynamics rather than creator-specific characteristics.
Finally, the study may be subject to selection bias. Only videos that were published and remained publicly available throughout the observation period were included in the analysis. Videos that were deleted or made private—potentially due to poor performance—are excluded, which may lead to an upward bias in estimated average engagement levels.

### Data Quality Issues

Despite thorough cleaning, some limitations remain:

- **Missing data:** Early videos (2019) have sparser data than recent videos
- **Metric definitions:** YouTube's internal definitions of "view" and "engagement" are proprietary and may have changed
- **Bot activity:** Some views, likes, or comments may stem from automated bots rather than genuine viewers

## Recommendations for Content Creators

Based on our empirical findings, we offer the following evidence-based recommendations:

**Recommendation 1: Optimize for Likes**

**Action:** Include explicit calls-to-action prompting viewers to like the video. Position these asks strategically:
- Early in the video (first 30 seconds) to capture immediate reactions
- After delivering high-value content moments to leverage reciprocity

**Evidence:** Like rate has the strongest correlation with overall engagement (r = 0.99). Even small increases in like percentage translate directly to higher engagement scores.

**Recommendation 2: Prioritize Engaged Niches Over Viral Reach**

**Action:** Resist the temptation to chase viral trends if they dilute brand identity or attract disengaged audiences. Focus content on core audience interests, even if this limits total view potential.

**Evidence:** Total views show weak correlation with engagement rate (r ≈ 0.20). Highly viewed videos do not necessarily perform better in engagement metrics.

**Recommendation 3: Maintain Retention, But Don't Overextend**

**Action:** Create content that holds viewer attention (strong hooks, pacing, editing) but avoid artificially lengthening videos beyond natural endpoints.

**Evidence:** Watch duration correlates moderately with engagement (r ≈ 0.30), but the relationship weakens for very long videos. Viewers may watch but not engage if content becomes repetitive.

**Recommendation 4: Test New Formats to Break the Plateau**

**Action:** Since engagement has remained stable for five years, experiment with:
- New content formats (shorts, live streams, collaborative videos)
- Different engagement mechanics (polls, pinned questions, challenges)
- Topic diversification within niche boundaries

**Evidence:** The flat temporal trend suggests existing strategies have reached optimization limits. Innovation may be necessary for growth.

**Recommendation 5: Reduce Emphasis on Posting Day Optimization**

**Action:** Maintain a consistent posting schedule (e.g., every Tuesday and Friday) rather than always targeting weekends. Prioritize content readiness over calendar optimization.

**Evidence:** Day-of-week effects are small (<0.7 percentage point difference). Operational simplicity and consistency may yield better long-term results than micromanaging posting times.

## Future Research Directions

There are several directions in which research can build on the findings presented in this report. One such direction involves the use of experimental or quasi-experimental designs, such as A/B testing, to evaluate the causal effects of different content formats (e.g., thumbnail images or calls-to-action) on varying levels of user engagement. A second avenue for future research would be to extend the scope of the analysis to additional creators across different channels, allowing the applicability of the findings to be assessed across a broader range of influencers.
Another promising direction involves the inclusion of qualitative measures of user engagement, for example through sentiment analysis of user comments using natural language processing techniques. Further research into algorithmic effects, particularly YouTube’s recommendation system, could also help clarify how platform algorithms mediate the relationship between content features and user engagement. Finally, longitudinal analyses of user behaviour could provide deeper insights into how engagement evolves over time and how viewers adapt to increased exposure to a specific creator or channel.

\newpage

# Conclusion

The aim of this report was to examine a seemingly simple question: what drives engagement on YouTube? By analysing 234,889 daily observations over a five-year period, several patterns emerged that challenge common assumptions and offer practical insights for content creators.
One of the most important findings is that like rate, rather than view count, plays the central role in driving engagement. This shifts the focus away from sheer popularity toward active audience interaction. The results show that viral reach does not necessarily lead to engaged audiences, that the day of publication has less influence than often assumed, and that engagement levels have remained largely stable over time despite consistent content quality.
These results also have clear practical implications. Rather than focusing primarily on maximising views or strictly optimising posting schedules, creators may benefit more from producing content that resonates with their core audience and explicitly encourages interaction. Even small increases in like rates can lead to noticeable improvements in overall engagement.
In addition to these practical insights, the report also contributes from an academic perspective. By providing a longitudinal and quantitative analysis of engagement dynamics on YouTube, it adds to existing research on social media analytics. The combination of data cleaning, correlation analysis, and visual exploration offers a structured and reproducible approach that can be applied to other digital platforms.
Looking ahead, engagement will continue to play a key role for creators, marketers, and platforms as the influencer economy grows. The findings of this report provide evidence-based guidance for improving engagement, while also highlighting that digital audience behaviour remains complex and strongly dependent on context.

\newpage

# References

::: {#refs}
:::

<!-- Note: Add references to a references.bib file -->
<!-- Example references that could be added:

@article{example2022,
  title={YouTube Algorithm and Engagement},
  author={Author, A. and Author, B.},
  journal={Journal of Digital Media},
  year={2022}
}

-->

\newpage

# Appendix: Code Repository and Reproducibility

All data cleaning, analysis, and visualization code is embedded within this Quarto document. The complete workflow is reproducible given access to the raw YouTube analytics data.

## Data Files

- `data/all_youtube_analytics.csv` - Raw export from YouTube Analytics
- `data/youtube_clean.csv` - Cleaned daily-level data
- `data/video_performance.csv` - Video-level aggregations
- `data/daily_performance.csv` - Date-level aggregations  
- `data/weekly_pattern.csv` - Day-of-week aggregations

## Author Contributions

**Athina Agiakatsikas:** Data cleaning and preparation, correlation analysis, manuscript writing (Methods and Results sections)

**Stella Christou:** Exploratory data analysis, visualization, manuscript writing (Introduction and Discussion sections)

Both authors contributed equally to the conceptual design, interpretation of findings, and final editing of the report.


